Real-space computation of $E$/$B$-mode maps I: Formalism, Compact Kernels, and Polarized Filaments

### 2018-07-31 articles/formalism/sections/real_space_formalism.tex
removed text change highlights and added acknowledgements
Added some clarifications based on Mathieu comments.

The same rotation $\mathcal{R}_{\hat{z}_q}(\psi)$ leaves the Euler angle $|\beta_{qe}|$ unaltered (it measures the angular distance between the points).  Thus we further conclude that the contribution to $(E+iB)(\hat n_e)$ from the position $n_q$ must have the generic form:
\beq
{}_{+2}X(\hat{n}_q) f(\beta_{qe})  e^{-i2\alpha_{qe}}
\eeq
for some real function $f$. \revisit{The total $(E+iB)$ field results from summing over all  contributions  from Stokes parameters ${}_{+2}X(\hat{n}_q)$ across the sphere.}

@@

The $E$/$B$ contribution from the Stokes parameters at some location $\hat{n}_q$ is given by the following expression (\eq{eq:op_qu2eb_rad} and \eq{eq:qu2eb_op}):
%
\beq  \label{eq:qu2eb_radiation_explicit}
\bar{S}_q(\hat n_e) = \bmat E_e \\ B_e  \emat_{q} =- {{}_{\mm}f}(\beta_{qe},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \alpha_{q e}) & \sin(2\alpha_{q e})\\  -\sin(2 \alpha_{q e})  & \cos(2 \alpha_{q e}) \emat  \bmat Q_{q} \\ U_{q}  \emat \Delta \Omega \,.
\eeq
%
The total map can be simply evaluated by summing over the contribution from the Stokes parameters at each location $\hat{n}_q$: $\bar{S} = \sum_{q=1}^{N_{\rm pix}} \bar{S}_q$. This operation can be cast concisely as:
%
\begin{subequations} \label{eq:qu2eb_radiation_concise}
\beqry 
\left[E + iB\right](\hat{n}_e) &=& -\Delta \Omega  \sum_{q=1}^{N_{\rm pix}} \Big[ {}_{+2}X(\hat{n}_{q}) e^{-i2\alpha_{q e}} \Big]  {{}_{\mm}f}(\beta_{q e}) \,, \\
&=& \Delta \Omega \sum_{q=1}^{N_{\rm pix}} {}_{+2}X(\hat{n}_{q}) \,  \mathcal{M}_{G}(\hat n_q) \,,
\eeqry
\end{subequations}
%
where $\Delta \Omega$ denotes the pixel area and the last line is a simple scalar multiplication between complex numbers.  The radiation kernel is then: $ \mathcal{M}_{G} =- \mm(\beta_{qe},\alpha_{qe})^*$ which can be thought of as the Green's function of the operator, since $[E +iB]= \mathcal{M}_{G}$ is the spin-0 scalar field generated from the delta-function Stokes field $[Q+iU] = [\delta(\hat{n}-\hat{n}_q)/{\Delta \Omega} + i0]$. We display the kernel later in \fig{fig:vis_kernel}.

@@

\paragraph{Convolution kernel.} We can also formulate the real space operator as a convolution operation, where the scalar field at $\hat n_e$ gathers contributions from the Stokes fields.  This is based around the inverse rotation from the previous section (to align the coordinate system at $\hat{n}_e$ with that at $\hat{n}_q$).  The inverse rotation Euler angles relates to the forward rotation Euler angles by the following relations: $\alpha_{eq}=-\gamma_{qe}$, $\beta_{eq} = -\beta_{qe}$ and  $\gamma_{eq} =-\alpha_{qe}$. Since the kernel depends on the cosine of the Euler angle $\beta$, it is immune to changes in its sign. The operator equation can be expressed as a function of the Euler angle $\gamma_{eq}$ as follows:
%
\beq \label{eq:qu2eb_convolution_explicit}
\bmat E_e \\ B_e  \emat =- \Delta \Omega\sum_{q=1}^{N_{\rm pix}}{{}_{\mm}f}(\beta_{eq},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \gamma_{eq}) & -\sin(2\gamma_{eq})\\  \sin(2 \gamma_{eq})  & \cos(2 \gamma_{eq}) \emat  \bmat Q_q \\ U_q  \emat \,,
\eeq
%
This formulation of the real space operator can be interpreted as integrating at some fixed location $\hat{n}_e$ the $E/B$ mode contribution arising from the Stokes parameters at all location $\hat{n}_q$ on the sphere. This operation can be expressed more concisely as follows:
%
\begin{subequations} \label{q:qu2eb_convolution_concise}
\beqry 
[E + iB](\hat{n}_e) &=& - \Delta \Omega \sum_{q=1}^{N_{\rm pix}}{{}_{\mm}f}(\beta_{eq},\ell_{\rm min},\ell_{\rm max}) {\Bigg( e^{i2 \gamma_{eq}}   {}_{+2}X (\hat{n}_q) \Bigg)}, \label{eq:qu2eb_physical}\\
&=& \Bigg\lbrace \mathcal{M}_{B} \star {}_{+2}X \Bigg\rbrace(\hat{n}_e) \,, \label{eq:qu2eb_convolution} 
\eeqry
\end{subequations}
%
where $\star$ denotes a convolution and $\mathcal{M}_{B} = -\mm(\beta_{eq},\gamma_{eq})$.  When $\mm$ is expressed as a function of the Euler angle $\gamma_{eq}$ it can be thought of as an effective instrument beam pointing to the direction $\hat{n}_e$. This many-to-one mapping from the spin-2 Stokes field on the sphere to the complex spin-0 (scalar) field at a point on the sphere is graphically represented in \fig{fig:planar_euler_angles}. We display the kernel in \fig{fig:vis_kernel}.
%--------------------------------------------------------

@@

The inverse operator expressed in terms of the function $\mm$ given in \eq{eq:qu2eb_gen_kernel} is given by the following equation:
%
\beq
{\bar O}^{-1}=- \Delta \Omega\bmat \mathcal{M}_{r} & -\mathcal{M}_{i} \\  \mathcal{M}_{i}  & \mathcal{M}_{r} \emat=- \Delta \Omega{{}_{\mm}f}(\beta_{eq},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \alpha_{qe}) & -\sin(2\alpha_{qe})\\  \sin(2 \alpha_{qe})  & \cos(2 \alpha_{qe}) \emat \,,
\eeq
%
where all the symbols have the same meaning as discussed in \sec{sec:qu2eb}. Note that the kernel in the above equation differs from the one in \eq{eq:op_qu2eb_rad} by a change in sign on the off-diagonals of the block matrix. When expressed in terms of the same set of Euler angles used to define the operator $\bar{O}$, it can be shown that the different forms of the real space operator are given by the following equations:
%
\beqry
    {}_{+2}X(\hat{n}_q) &&=  \Delta \Omega \sum_{e=1}^{N_{\rm pix}} [E+iB](\hat{n}_{e})\   \mathcal{M}^*_{B}(\hat{n}_e) \hspace{0.8cm}\textrm{\emph{Radiation kernel}},\label{eq:eb2qu_radiation} \\
    {}_{+2}X(\hat{n}_q) &&= \Bigg\lbrace \mathcal{M}^*_{G} \star [E+iB] \Bigg\rbrace(\hat{n}_q) \hspace{1.4cm}\textrm{\emph {Convolution kernel}}, \label{eq:eb2qu_convolution}
\eeqry
%
where all the symbols have the same meaning as defined in \sec{sec:qu2eb}. Note that the conjugated forms of the radiation kernel (Green's function) and the convolution kernel (effective beam) for the operator $\bar{O}$ have their roles reversed for the inverse operator $\bar{O}^{-1}$.
%--------------------------------------------------------

@@

  \caption{Real and imaginary parts of real space kernels for $Q/U$ to $E/B$ translation (and vice versa).  The function $\mathcal{M}_{G}$ is the Green's function (radiation kernel) that gives $E+iB$ for a delta function Stokes input $_{+2}X = Q+iU$ at the center.  On the other hand, $\mathcal{M}^*_{B}$ is the Green's function (radiation kernel) that gives $_{+2}X$ for a delta function scalar input $E+iB$ at the center.  The black circles denotes the position of the center around which the kernels have been evaluated while the black star marks the location of the North Pole. The four rows depict the kernels at different latitudes on the sphere.   The kernels are unchanged for center points at constant latitude.  $\mathcal{M}_G$ is invariant over the sphere because $E/B$ fields are coordinate independent.  In the convolution kernels, $\mathcal{M}_G$ and $\mathcal{M}_B$ are conjugated and switch roles.  The kernels have been evaluated with the band limit $\ell \in [2,192]$ and sampled at the HEALPix resolution parameter $N_{\rm side}=2048$. Each panel is approximately $16^{\circ} \times 16^{\circ}$ in size with grid lines every 2 degrees. } \label{fig:vis_kernel} 

@@

The kernel $\mathcal{M}_G$ is the Green's function of the operator that transforms Stoke parameters to coordinate independent $E/B$.  The coordinate independence implies that the real and imaginary parts of the kernel do not vary with changes in the galactic latitude and longitude of the central pixel. In particular these functions are not distorted when a part of the domain overlaps with the poles, as can be seen in the first two rows of \fig{fig:vis_kernel}. From \eq{eq:qu2eb_radiation_concise}, we can see the $E/B$ patterns that Kronecker $\delta$-functions in the Stokes parameter pixels produce:
%
\beq
\bmat E= +\textrm{Re}(\mathcal{M}_G) \\ B =+\textrm{Im}(\mathcal{M}_G)  \emat  \leftarrow \bmat Q=\frac{\delta_{\hat{n},\hat{n}_q}}{\Delta \Omega}\\ U=0 \emat ;\qquad
\bmat E= -\textrm{Im}(\mathcal{M}_G) \\ B = + \textrm{Re}(\mathcal{M}_G)  \emat  \leftarrow \bmat Q=0 \\ U=\frac{\delta_{\hat{n},\hat{n}_q}}{\Delta \Omega} \emat.
\eeq
%
On the other hand, $\mathcal{M}^*_B$ is the Green's function of the operator that transforms $E/B$ to coordinate-dependent Stokes parameters $Q/U$.   The kernel $\mathcal{M}^*_B$ does not change with the central longitude, but varies as a function of galactic latitude. This latitude-dependent shape carries the  coordinate dependence of the Stokes parameters. From \eq{eq:eb2qu_radiation}, we $\mathcal{M}^*_B$ arises from $\delta$-functions in $E/B$:
%
\beq
\bmat Q= +\textrm{Re}(\mathcal{M}^*_B) \\ U = +\textrm{Im}(\mathcal{M}^*_B)  \emat  \leftarrow \bmat E=\frac{\delta_{\hat{n},\hat{n}_e}}{\Delta \Omega}\\ B=0 \emat;\qquad
\bmat Q= -\textrm{Im}(\mathcal{M}^*_B) \\ U = +\textrm{Re}(\mathcal{M}^*_B)  \emat  \leftarrow \bmat E=0 \\ B=\frac{\delta_{\hat{n},\hat{n}_e}}{\Delta \Omega} \emat\,.
\eeq
%
Recall that the complex conjugates of these functions switch roles to form the convolution kernels. (For example, $\mathcal{M}_G^*$ is the convolution kernel for $E/B \rightarrow Q/U$.)  It is also interesting to note that the radiation kernel ($\mm_{G}$) and the convolution kernel ($\mm_{B}$) become progressively identical as one approaches the equator as seen in  \fig{fig:vis_kernel}, which is a consequence of $\gamma \simeq -\alpha$ in the vicinity of the equator. The equator is the place on the sphere most akin to the flat sky case, where the radiation and convolution kernels are expected to be identical.

>>>>

The same rotation $\mathcal{R}_{\hat{z}_q}(\psi)$ leaves the Euler angle $|\beta_{qe}|$ unaltered (it measures the angular distance between the points).  Thus we further conclude that the contribution to $(E+iB)(\hat n_e)$ from the position $n_q$ must have the generic form:
\beq
{}_{+2}X(\hat{n}_q) f(\beta_{qe})  e^{-i2\alpha_{qe}}
\eeq
for some real function $f$. The total $(E+iB)$ field results from summing over all  contributions  from Stokes parameters ${}_{+2}X(\hat{n}_q)$ across the sphere.

@@

The $E$/$B$ contribution from the Stokes parameters at some location $\hat{n}_q$ is given by the following expression (\eq{eq:op_qu2eb_rad} and \eq{eq:qu2eb_op}):
%
\beq  \label{eq:qu2eb_radiation_explicit}
\bar{S}_q(\hat n_e) = \bmat E_e \\ B_e  \emat_{q} =- {{}_{\mm}f}(\beta_{qe},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \alpha_{q e}) & \sin(2\alpha_{q e})\\  -\sin(2 \alpha_{q e})  & \cos(2 \alpha_{q e}) \emat  \bmat Q_{q} \\ U_{q}  \emat \Delta \Omega \,.
\eeq
%
The total map can be simply evaluated by summing over the contribution from the Stokes parameters at each location $\hat{n}_q$: $\bar{S} = \sum_{q=1}^{N_{\rm pix}} \bar{S}_q$. This operation can be cast concisely as:
%
\begin{subequations} \label{eq:qu2eb_radiation_concise}
\beqry 
\left[E + iB\right](\hat{n}_e) &=& -\Delta \Omega  \sum_{q=1}^{N_{\rm pix}} \Big[ {}_{+2}X(\hat{n}_{q}) e^{-i2\alpha_{q e}} \Big]  {{}_{\mm}f}(\beta_{q e}) \,, \\
&=& \Delta \Omega \sum_{q=1}^{N_{\rm pix}} {}_{+2}X(\hat{n}_{q}) \,  \mathcal{M}_{G}(\hat n_q) \,,
\eeqry
\end{subequations}
%
where $\Delta \Omega$ denotes the pixel area and the last line is a simple scalar multiplication between complex numbers.  The radiation kernel is then: $ \mathcal{M}_{G} =- \mm(\beta_{qe},\alpha_{qe})^*$ which can be thought of as the Green's function of the operator, since $[E +iB]= \mathcal{M}_{G}$ is the spin-0 scalar field generated from the delta-function Stokes field $[Q+iU] = [\delta(\hat{n},\hat{n}_q)/{\Delta \Omega} + i0]$. We display the kernel later in \fig{fig:vis_kernel}.

@@

\paragraph{Convolution kernel.} We can also formulate the real space operator as a convolution operation, where the scalar field at $\hat n_e$ gathers contributions from the Stokes fields.  This is based around the inverse rotation from the previous section (to align the coordinate system at $\hat{n}_e$ with that at $\hat{n}_q$).  The inverse rotation Euler angles relates to the forward rotation Euler angles by the following relations: $\alpha_{eq}=-\gamma_{qe}$, $\beta_{eq} = -\beta_{qe}$ and  $\gamma_{eq} =-\alpha_{qe}$. Since the kernel depends on the cosine of the Euler angle $\beta$, it is immune to changes in its sign. The operator equation can be expressed as a function of the Euler angle $\gamma_{eq}$ as follows:
%
\beq \label{eq:qu2eb_convolution_explicit}
\bmat E_e \\ B_e  \emat =- \Delta \Omega\sum_{q=1}^{N_{\rm pix}}{{}_{\mm}f}(\beta_{eq},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \gamma_{eq}) & -\sin(2\gamma_{eq})\\  \sin(2 \gamma_{eq})  & \cos(2 \gamma_{eq}) \emat  \bmat Q_q \\ U_q  \emat \,,
\eeq
%
This formulation of the real space operator can be interpreted as integrating at some fixed location $\hat{n}_e$ the $E/B$ mode contribution arising from the Stokes parameters at all location $\hat{n}_q$ on the sphere. This operation can be expressed more concisely as follows:
%
\begin{subequations} \label{q:qu2eb_convolution_concise}
\beqry 
[E + iB](\hat{n}_e) &=& - \Delta \Omega \sum_{q=1}^{N_{\rm pix}}{{}_{\mm}f}(\beta_{eq},\ell_{\rm min},\ell_{\rm max}) {\Bigg( e^{i2 \gamma_{eq}}   {}_{+2}X (\hat{n}_q) \Bigg)}, \label{eq:qu2eb_physical}\\
&=& \Bigg\lbrace \mathcal{M}_{B} \star {}_{+2}X \Bigg\rbrace(\hat{n}_e) \,, \label{eq:qu2eb_convolution} 
\eeqry
\end{subequations}
%
where $\star$ denotes a convolution and $\mathcal{M}_{B} = -\mm(\beta_{eq},\gamma_{eq})$.  When $\mm$ is expressed as a function of the Euler angle $\gamma_{eq}$ it can be thought of as an effective instrument beam pointing to the direction $\hat{n}_e$. This many-to-one mapping from the spin-2 Stokes field on the sphere to the complex spin-0 (scalar) field at a point on the sphere is graphically represented in \fig{fig:planar_euler_angles}. \revisit{We display the complex conjugate of this kernel in \fig{fig:vis_kernel}.  (Later we see that the complex conjugate of this kernel is the radiation kernel for the inverse transform.)}
%--------------------------------------------------------

@@

The inverse operator expressed in terms of the function $\mm$ given in \eq{eq:qu2eb_gen_kernel} is given by the following equation:
%
\beq
{\bar O}^{-1}=- \Delta \Omega\bmat \mathcal{M}_{r} & -\mathcal{M}_{i} \\  \mathcal{M}_{i}  & \mathcal{M}_{r} \emat=- \Delta \Omega{{}_{\mm}f}(\beta_{eq},\ell_{\rm min},\ell_{\rm max})\bmat \cos(2 \alpha_{qe}) & -\sin(2\alpha_{qe})\\  \sin(2 \alpha_{qe})  & \cos(2 \alpha_{qe}) \emat \,,
\eeq
%
where all the symbols have the same meaning as discussed in \sec{sec:qu2eb}. Note that the kernel in the above equation differs from the one in \eq{eq:op_qu2eb_rad} by a change in sign on the off-diagonals of the block matrix. When expressed in terms of the same set of Euler angles used to define the operator $\bar{O}$, it can be shown that the different forms of the real space operator are given by the following equations:
%
\beqry
    {}_{+2}X(\hat{n}_q) &&=  \Delta \Omega \sum_{e=1}^{N_{\rm pix}} [E+iB](\hat{n}_{e})\   \mathcal{M}^*_{B}(\hat{n}_e) \hspace{0.8cm}\textrm{\emph{Radiation kernel}},\label{eq:eb2qu_radiation} \\
    {}_{+2}X(\hat{n}_q) &&= \Bigg\lbrace \mathcal{M}^*_{G} \star [E+iB] \Bigg\rbrace(\hat{n}_q) \hspace{1.4cm}\textrm{\emph {Convolution kernel}}, \label{eq:eb2qu_convolution}
\eeqry
%
where all the symbols have the same meaning as defined in \sec{sec:qu2eb}.  \revisit{Note that $M_B$ (the convolution kernel before for $Q/U \rightarrow E/B$) is here the radiation kernel (as $M_B^*$) for $E/B \rightarrow Q/U$.  Similarly $M_G$ (the radiation kernel before for $Q/U \rightarrow E/B$) is here the convolution kernel (as $M_G^*$) for $E/B \rightarrow Q/U$.}
Thus the conjugated forms of the radiation kernel (Green's function) and the convolution kernel (effective beam) for the operator $\bar{O}$ reverse roles for the inverse operator $\bar{O}^{-1}$.
%--------------------------------------------------------

@@

\caption{Real and imaginary parts of real space kernels for $Q/U$ to $E/B$ translation (and vice versa).  The function $\mathcal{M}_{G}$ is the Green's function (radiation kernel) that gives $E+iB$ for a delta function Stokes input $_{+2}X = Q+iU$ at the center.  On the other hand, $\mathcal{M}^*_{B}$ is the Green's function (radiation kernel) that gives $_{+2}X$ for a delta function scalar input $E+iB$ at the center.  The black circles denotes the position of the center around which the kernels have been evaluated while the black star marks the location of the North Pole. The four rows depict the kernels at different latitudes on the sphere.   The kernels are unchanged for center points at constant latitude.  $\mathcal{M}_G$ is invariant over the sphere because $E/B$ fields are coordinate independent.
\revisit{The \textit{convolution kernels} for these translations are respectively $\mathcal{M}_B$ and $\mathcal{M}_G^*$: the functions are conjugated and switch roles between forward and inverse translation.}  The kernels have been evaluated with the band limit $\ell \in [2,192]$ and sampled at the HEALPix resolution parameter $N_{\rm side}=2048$. Each panel is approximately $16^{\circ} \times 16^{\circ}$ in size with grid lines every 2 degrees. } \label{fig:vis_kernel} 

@@

The kernel $\mathcal{M}_G$ is the Green's function of the operator that transforms Stoke parameters to coordinate independent $E/B$.  The coordinate independence implies that the real and imaginary parts of the kernel do not vary with changes in the galactic latitude and longitude of the central pixel. In particular these functions are not distorted when a part of the domain overlaps with the poles, as can be seen in the first two rows of \fig{fig:vis_kernel}. From \eq{eq:qu2eb_radiation_concise}, we can see the $E/B$ patterns that Kronecker $\delta$-functions in the Stokes parameter pixels produce:
%
\beq
\bmat E= +\textrm{Re}(\mathcal{M}_G) \\ B =+\textrm{Im}(\mathcal{M}_G)  \emat  \leftarrow \bmat Q=\frac{\delta_{\hat{n},\hat{n}_q}}{\Delta \Omega}\\ U=0 \emat ;\qquad
\bmat E= -\textrm{Im}(\mathcal{M}_G) \\ B = + \textrm{Re}(\mathcal{M}_G)  \emat  \leftarrow \bmat Q=0 \\ U=\frac{\delta_{\hat{n},\hat{n}_q}}{\Delta \Omega} \emat.
\eeq
%
On the other hand, $\mathcal{M}^*_B$ is the Green's function of the operator that transforms $E/B$ to coordinate-dependent Stokes parameters $Q/U$.   The kernel $\mathcal{M}^*_B$ does not change with the central longitude, but varies as a function of galactic latitude. This latitude-dependent shape carries the  coordinate dependence of the Stokes parameters. From \eq{eq:eb2qu_radiation}, we $\mathcal{M}^*_B$ arises from $\delta$-functions in $E/B$:
%
\beq
\bmat Q= +\textrm{Re}(\mathcal{M}^*_B) \\ U = +\textrm{Im}(\mathcal{M}^*_B)  \emat  \leftarrow \bmat E=\frac{\delta_{\hat{n},\hat{n}_e}}{\Delta \Omega}\\ B=0 \emat;\qquad
\bmat Q= -\textrm{Im}(\mathcal{M}^*_B) \\ U = +\textrm{Re}(\mathcal{M}^*_B)  \emat  \leftarrow \bmat E=0 \\ B=\frac{\delta_{\hat{n},\hat{n}_e}}{\Delta \Omega} \emat\,.
\eeq
%
Recall that the complex conjugates of these functions switch roles to form the convolution kernels. (For example, $\mathcal{M}_G^*$ is the convolution kernel for $E/B \rightarrow Q/U$.)  It is also interesting to note that the radiation kernel ($\mm_{G}$) and the convolution kernel ($\mm_{B}$) become progressively identical as one approaches the equator, which is a consequence of $\gamma \simeq -\alpha$ in the vicinity of the equator.  \revisit{This is visible in \fig{fig:vis_kernel} (although it displays $M_B^*$ not $M_B$).}  The equator is the place on the sphere most akin to the flat sky case, where the radiation and convolution kernels are expected to be identical.
